---
title: "blastn taxonomic curation"
author: "Kimberly Ledger"
date: "2025-12-18"
output: html_document
---

taxonomic id of blastn output. this code is useful for assigning higher classifications (genus, family, etc) for ASVs that have multispecies hits.   

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-libraries}
library(dplyr)
library(tidyr)
library(stringr)
library(tidyverse)
library(here)
```


## load output from blastn 
```{r load-output-from-taxonkit}
taxonomy <-read.delim(here("dada2_blastn_outputs", "blastnresults"), header = FALSE, na.strings=c(""," ","NA"))
head(taxonomy)
colnames(taxonomy) <- c("ASV", "subject", "pident", "length", "mismatch", "gapopen",
  "qstart", "qend", "sstart", "send", "evalue", "bitscore", "qseq", "sseq")
```

## clean up the header and seperate subject 
```{r}
# use the full taxonomy rather than the seq id to collapse identical entries
tax_df <- taxonomy %>%
  filter(length > 100) %>% # make sure all retained matches are >100 bp
  select(ASV, subject, pident, length) %>%
  separate(subject, into = c("subject_id", "taxonomy"), sep = "###") %>%
  # Split taxonomy string into ranks
  separate(taxonomy, into = c("root","kingdom","phylum","class","order","family","genus","species"),
           sep = ";", fill = "right") %>%
  select(!root) %>% 
  # Separate name and taxID for each rank
  mutate(across(c(kingdom, phylum, class, order, family, genus),
                ~ str_split_fixed(., "_", 2))) %>% 
  mutate(species = str_match(species, "^(.*_.*)_(\\d+)$")[,2:3]) %>% 
  mutate(kingdom = kingdom[,1],
    phylum  = phylum[,1],
    class   = class[,1],
    order   = order[,1],
    family  = family[,1],
    genus   = genus[,1],
    species = species[,1]) %>%
  select(ASV, pident, length, kingdom, phylum, class, order, family, genus, species) %>%
  unique() %>%
  filter(!str_detect(species, "_x_")) %>% ##remove hybrids
  mutate(species = str_replace_all(species, "_", " "))   
```

## now look at the fish and figure out what fish taxa are in/out of our range
```{r}
to_check_range <- tax_df %>%
  filter(class == "Actinopteri") %>%
  ungroup() %>%
  select(species, genus, family, order, class, phylum) %>%
  unique()
```

## check ranges for species using rfishbase 
```{r}
#library(remotes)
#remotes::install_github("ropensci/rfishbase")
library(rfishbase)

#one-time download of all fishbase tables... this takes a bit 
#fb_import()

#first, validate species names using rfishbase synonyms
spp_df <- synonyms(to_check_range$species)

#check the synonyms
syn <- spp_df %>% 
  filter(Status == "synonym")

#rename any synonyms with the validated name  
to_check_range_validated <- to_check_range %>%
  rename(synonym = species) %>%
  left_join(syn, by = "synonym") %>%
  mutate(synonym = ifelse(is.na(Species), synonym, Species)) %>%
  select(genus:synonym) %>%
  rename(species = synonym) %>%
  unique()

#get distribution info 
spp_distribution <- ecosystem(to_check_range_validated$species) %>%
  select(Species, EcosystemName) %>%
  unique()

#add column to designate if we will consider a species as "in range"- for this study 
spp_distribution_range <- spp_distribution %>%
  mutate(in_range = ifelse(is.na(EcosystemName), NA, "no"),
         in_range = ifelse(EcosystemName == "Pacific Ocean", "yes", in_range),
         in_range = ifelse(EcosystemName == "East Bering Sea", "yes", in_range))

in_range_df <- spp_distribution_range %>%
  filter(in_range == "yes") %>%
  arrange(Species)

out_range_df <- spp_distribution_range %>%
  filter(in_range != "yes") %>%
  select(Species) %>% unique() %>%
  filter(!Species %in% in_range_df$Species)
```

## put all non-fish in their own table
```{r}
not_Actinopteri <- tax_df %>%
  filter(class != "Actinopteri") 
```

## keep only the "in range" fish species in the taxon dataframe - and all non-fish - and the positive control species (Acipenser fulvescens)
```{r}
in_range_spp <- unique(in_range_df$Species)

taxa_keepers <- c(in_range_spp, unique(not_Actinopteri$species),  "Acipenser fulvescens")

taxon_df <- tax_df %>%
  filter(species %in% taxa_keepers)

#syn %>% filter(Species %in% taxa_keepers)

#rename synonyms with validated names 
taxon_df <- taxon_df %>%
  mutate(species = ifelse(species == "Gadus ogac", "Gadus macrocephalus", species),
         species = ifelse(species == "Cleisthenes herzensteini", "Cleisthenes pinetorum", species),
         species = ifelse(species == "Kareius bicoloratus", "Platichthys bicoloratus", species), 
         genus = ifelse(species == "Kareius bicoloratus", "Platichthys", genus),
         species = ifelse(species == "Sebastes matsubarae", "Sebastes matsubarai", species),
         species = ifelse(species == "Helicolenus hilgendorfi", "Helicolenus hilgendorfii", species), 
         species = ifelse(species == "Embassichthys bathybius", "Microstomus bathybius", species), 
         genus = ifelse(species == "Microstomus bathybius", "Microstomus", genus), 
         species = ifelse(species == "Zestichthys tanakae", "Bothrocara tanakae", species),
         genus = ifelse(species == "Zestichthys tanakae", "Bothrocara", genus),
         species = ifelse(species == "Hypsagonus jordani", "Agonomalus jordani", species),
         genus = ifelse(species == "Hypsagonus jordani", "Agonomalus", genus),
         species = ifelse(species == "Ulcina olrikii", "Aspidophoroides olrikii", species),
         genus = ifelse(species == "Ulcina olrikii", "spidophoroides", genus),
         species = ifelse(species == "Gasterosteus gymnurus", "Gasterosteus aculeatus", species))
```

## remove a few more species that missed in the geo filter b/c they occur in NW Pacific, etc  
```{r}
remove <- c("Acanthopsetta nadeshnyi", "Cleisthenes pinetorum", "Hippoglossoides dubius", "Lepidopsetta mochigarei", "Lycodes fasciatus", "Lycodes nakamurae", "Myoxocephalus brandtii", "Megalocottus platycephalus", "Liopsetta glacialis", "Liopsetta pinnifasciata", "Icelus cataphractus", "Glyptocephalus stelleri", "Sebastiscus marmoratus") 

taxon_df <- taxon_df %>%
  filter(!species %in% remove)
```

## figure out what ASVs only match a single species and assign taxonomic based on %ID
```{r}
single_spp_seqs <- taxon_df %>% 
  group_by(ASV) %>%
  add_tally(name = "n_taxa") %>%
  filter(n_taxa == 1)

#modify headers to match with filtering to follow 
single_spp_seqs <- single_spp_seqs %>%
  rename(seq_percID = pident) %>%
  mutate(id_rank = 1,
         top_perc = seq_percID,
         diff = 0) %>%
  select(!length) %>%
  select(!n_taxa)

single_spp_seqs_df <- single_spp_seqs %>%
  mutate(
    # 1. Species assignment (if > 98)
         taxon = ifelse(top_perc > 98, species, NA),
         taxonomic_level = ifelse(top_perc > 98, "species", NA),
         
    # 2. Genus assignment (if between 96.0 and 98.0)
         taxon = ifelse(is.na(taxon) & top_perc > 96 & top_perc <= 98, genus, taxon),
         taxonomic_level = ifelse(is.na(taxonomic_level) & top_perc > 96 & top_perc <= 98, "genus", taxonomic_level))
```

## remove the single-species seqs from the dataframe and then rank the hits by % identity for the remaining seqs
```{r}
seq_id_diff <- taxon_df %>%
  anti_join(., single_spp_seqs) %>%  ## comment out this line if there are no single species ASVs
  select(-length) %>%
  group_by(ASV, species, genus, family, order, class, phylum, kingdom) %>%
    mutate(seq_percID = max(pident)) %>%
    group_by(ASV, species, genus, family, order, class, phylum, kingdom, seq_percID) %>%
  summarise(max(seq_percID)) %>% # take just the top hit for each taxon (for each sequence)
  select(-`max(seq_percID)`) %>%
  ungroup() %>%
  group_by(ASV) %>%
      mutate(id_rank = rank(desc(seq_percID), ties.method = "min")) %>% # rank the taxonomic hits per sequence by % id
       mutate(top_perc = max(seq_percID)) %>% # designate the highest % id for the best taxonomic hit in each sequence (in some, but not all cases, this is 100%)   
      mutate(diff = top_perc - seq_percID) %>% # calculate the difference between the % identity of the top hit and each subsequent taxonomic hit
        arrange(diff)
```

Now you have the single best entry for each species for each sequence ranked and with the difference between the first and second ranked entries calculated.

For sequences with multiple top hits, 
(1) sequences with a top hit within 99.4%, consider all species within 0.5% of best
(2) sequences with a top hit less than 99.4%, consider all species within 1% of best

```{r}
to_remove_low_perc_hits_99.4 <- seq_id_diff %>%
  ungroup() %>%
  group_by(ASV) %>%
  filter(top_perc > 99.4) %>%
  filter(diff > 0.5)

to_remove_low_perc_hits_remaining <- seq_id_diff %>%
  ungroup() %>%
  group_by(ASV) %>%
  filter(top_perc <= 99.4) %>%
  filter(diff > 1)
```


## now remove those low percent hit matches 
```{r}
to_remove_low_perc_hits <- to_remove_low_perc_hits_99.4 %>%
   bind_rows(to_remove_low_perc_hits_remaining) 

keepers <- seq_id_diff %>%
  anti_join(to_remove_low_perc_hits)
```

## so now determine whether ASVs should be assigned to genus, family, order, etc. 
```{r}
singletons <- keepers %>%
  select(ASV) %>%
  tally() %>%
  filter(n == 1)

# these are the seqs that now have only a single match
singleton_df <- singletons %>%
  left_join(keepers) %>%
  select(-n) %>%
  mutate(
    # 1. Species assignment (if > 98)
         taxon = ifelse(top_perc > 98, species, NA),
         taxonomic_level = ifelse(top_perc > 98, "species", NA),
         
    # 2. Genus assignment (if between 96.0 and 98.0)
         taxon = ifelse(is.na(taxon) & top_perc > 96 & top_perc <= 98, genus, taxon),
         taxonomic_level = ifelse(is.na(taxonomic_level) & top_perc > 96 & top_perc <= 98, "genus", taxonomic_level)) %>%
  bind_rows(single_spp_seqs_df) # combine the single spp data


## Genus-level matches
# remove the singletons from the bigger df 
single_genus <- keepers %>%
  anti_join(singleton_df)%>% # at best, these should be genus-level matches
  group_by(ASV, genus) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single genus
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "genus") %>%
  mutate(taxon = genus)
  

## Family-level matches
single_family <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  group_by(ASV, family) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single family
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "family") %>%
  mutate(taxon = family)


## Order-level matches
single_order <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  group_by(ASV, order) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>%
  filter(n == 1) %>% # seqs that match a single order
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "order") %>%
  mutate(taxon = order)


## Class-level matches
single_class <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  group_by(ASV, class) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single class
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "class") %>%
  mutate(taxon = class)


## Phylum-level matches
single_phylum <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(ASV, phylum) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single phylum
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "phylum") %>%
  mutate(taxon = phylum)

# Kingdom level matches
single_kingdom <- keepers %>%
  anti_join(singleton_df)%>%
  anti_join(single_genus) %>%
  anti_join(single_family) %>%
  anti_join(single_order) %>%
  anti_join(single_class) %>%
  group_by(ASV, kingdom) %>%
  tally() %>%
  ungroup() %>%
  group_by(ASV) %>%
  tally() %>% 
  filter(n == 1) %>% # seqs that match a single kingdom
  select(-n) %>%
  left_join(., keepers) %>%
  mutate(taxonomic_level = "kingdom") %>%
  mutate(taxon = kingdom)
```

## recombine the full data set now that the appropriate level of taxonomy has been determined
```{r}
sorted_tax_df <- bind_rows(single_kingdom, single_phylum, single_class, single_order, single_family, single_genus, singleton_df)
```

## Create output taxonomy data frames
```{r}
uncollapsed_taxonomy <- sorted_tax_df %>%
  select(-top_perc, -id_rank) %>%
  unique() %>%
  arrange(ASV)
``` 

add in some additional manual curation:

```{r}
curated_taxonomy <- uncollapsed_taxonomy %>%
  mutate(taxon = ifelse(taxon == "Atheresthes evermanni", "Atheresthes", taxon), ##lump Atheresthes all together b/c few spp-level asv ids 
         taxonomic_level = ifelse(taxon == "Atheresthes", "genus", taxonomic_level), 
         taxon = ifelse(ASV == "ASV112", "Pleuronectes quadrituberculatus", taxon),
         taxonomic_level = ifelse(ASV == "ASV112", "species", taxonomic_level),
         taxon = ifelse(ASV == "ASV208", "Hemitripterus bolini", taxon),
         taxonomic_level = ifelse(ASV == "ASV208", "species", taxonomic_level)) %>%
  mutate(taxon = ifelse(ASV == "ASV181", "Careproctus", taxon),   ## all Careproctus to genus
         taxonomic_level = ifelse(ASV == "ASV181", "genus", taxonomic_level)) %>%
  mutate(taxon = ifelse(taxon == "Pleuronectidae", "Hippoglossoides", taxon),
         taxonomic_level = ifelse(taxon == "Hippoglossoides", "genus", taxonomic_level))  ## change Pleuronectidae to Hippoglossoides 

rows_to_remove_1 <- curated_taxonomy %>%
  filter(ASV == "ASV112" & species == "Platichthys stellatus") 
rows_to_remove_2 <- curated_taxonomy %>%
  filter(ASV == "ASV208" & species != "Hemitripterus bolini") 
rows_to_remove_3 <- curated_taxonomy %>%
  filter(taxon == "Hippoglossoides" & genus == "Limanda") 

curated_taxonomy_2 <- curated_taxonomy %>%
  anti_join(rows_to_remove_1) %>%
  anti_join(rows_to_remove_2) %>%
  anti_join(rows_to_remove_3)
```

notes:
Agonidae = Leptagonus decagonus or Sarritor frenatus
Pleuronectidae = Hippoglossoides elassodon, Hippoglossoides robustus, or Limanda sakhalinensis
Cottidae = a variety of different sculpin groupings 
Lycodes = mostly Lycodes brevipes but also some Lycodes palearis

```{r}
# and then collapse that down to just a single taxon per ASV
collapsed_taxonomy <- curated_taxonomy_2 %>%
  mutate(species = ifelse(taxonomic_level == "family", NA, species),
         genus = ifelse(taxonomic_level == "family", NA, genus),
         species = ifelse(taxonomic_level == "genus", NA, species),
         family = ifelse(taxonomic_level == "order", NA, family),
         genus = ifelse(taxonomic_level == "order", NA, genus),
         species = ifelse(taxonomic_level == "order", NA, species)) %>%
  select(ASV, taxon, taxonomic_level, species, genus, family, order, class, phylum, kingdom) %>%
  unique()
```

```{r}
write.csv(uncollapsed_taxonomy, here("tax_outputs", "uncollapsed_taxonomy.csv"))
write.csv(collapsed_taxonomy, here("tax_outputs", "collapsed_taxonomy.csv"))
```
