---
title: "metaprobe summary"
output: html_document
date: "2025-12-18"
---

libraries
```{r}
library(tidyverse)
library(here)
library(ggplot2)
#devtools::install_github("thomasp85/scico")
library(scico)
library(vegan)
library(reshape2)
```

load decontaminated ASV table 
```{r}
asv_table <- readRDS(here("decon_outputs", "decontaminated_asv_table.RDS"))
```

convert to taxon table
and remove any remaining pc reads 
```{r}
taxon_table <- asv_table %>%
  filter(taxon != "Acipenser fulvescens") %>% 
  group_by(extraction_ID, sample_type, sample_id, taxon, taxonomic_level, species, genus, family, order) %>%
  summarize(reads = sum(reads)) %>%
  filter(reads > 0)
```

load sample metadata 
```{r}
metadata <- read.csv(here("BeringSea2024_metaprobes_metadata.csv"))

# get depth from catch data 
catch <- read.csv(here("catch_outputs", "metaprobe_haul_catch.csv")) %>%
  select(haul, depth_m) %>% unique()

taxon_table_w_meta <- taxon_table %>% 
  left_join(metadata) %>%
  left_join(catch, by = "haul")
```

make a summary table of the species and reads - not including any field blanks in this 
```{r}
taxon_table_summary <- taxon_table_w_meta %>%
  filter(sample_type == "sample") %>%
  group_by(taxon, species, genus, family, order) %>%
  summarize(total_reads = sum(reads),
            n_tech = n_distinct(sample_id),
            n_extract = n_distinct(extraction_ID),
            n_probes = n_distinct(alternative_ID)) %>%
  arrange(desc(total_reads))
```

remove a few assignments that have few reads and are at poor taxonomic resolution (probably sequencing errors from species that are actually identified)
```{r}
## check if family level assignments are always with species/genus assignmetns...
samples_w_Cottidae <- taxon_table_w_meta %>% 
  filter(taxon == "Cottidae") 

taxon_table_Cottidae <- taxon_table_w_meta %>% 
  filter(sample_id %in% samples_w_Cottidae$sample_id)

## filter
taxon_table_filtered <- taxon_table_w_meta %>%
  filter(!taxon %in% c("Lycodes", "Cottidae"))
```


```{r}
taxon_table_summary <- taxon_table_filtered %>%
  filter(sample_type == "sample") %>%
  group_by(taxon, species, genus, family, order) %>%
  summarize(total_reads = sum(reads),
            n_tech = n_distinct(sample_id),
            n_extract = n_distinct(extraction_ID),
            n_probes = n_distinct(alternative_ID)) %>%
  arrange(desc(total_reads))

#write.csv(taxon_table_summary, here("metaprobe_outputs", "taxon_table_summary.csv"), row.names = F)
```

```{r}
taxon_table_filtered %>%
  group_by(haul) %>%
  summarize(n_taxa = n_distinct(taxon)) %>%
  ungroup() %>%
  summarize(min = min(n_taxa),
            mean = mean(n_taxa),
            max = max(n_taxa))
```

take a closer look at gadus... 
```{r}
gadus <- taxon_table_filtered %>%
  filter(genus == "Gadus") %>%
  group_by(taxon) %>%
  summarise(total_reads = sum(reads))
gadus

#prop of read to walleye pollock 
4058893/(sum(gadus$total_reads))
#prop of read to pcod
12714/(sum(gadus$total_reads))
#prop of read to gadus
127482/(sum(gadus$total_reads))

## catch weight from associated hauls was 
# 18078.545 kg pollock 
18078.545 / (18078.545+917.214)
# 917.214 kg pcod 
917.214 / (18078.545+917.214)
```


lump the gadus reads
```{r}
taxon_table_2 <- taxon_table_filtered %>%
  mutate(taxon = ifelse(taxon == "Gadus chalcogrammus", "Gadus", taxon),
         taxon = ifelse(taxon == "Gadus macrocephalus", "Gadus", taxon),
         taxonomic_level = ifelse(taxon == "Gadus", "genus", taxonomic_level),
         species = ifelse(taxon == "Gadus", NA, species)) %>%
  group_by(extraction_ID, sample_type, sample_id, taxon, taxonomic_level, species, genus, family, order, alternative_ID, haul, net, depth_m) %>%
  summarize(reads = sum(reads))
```



just start by comparing the samples/pcr-reps from the same haul 
```{r}
scico_pal <- scico(16, palette = 'roma') ## may need to change pal# depending on haul being plotted 

taxon_table_2 %>%
  filter(haul == 190) %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ggplot(aes(x=sample_id, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~net, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "sample ID") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )
```


okay, next step will need to be taking the mean of the extraction/pcr replicates for samples.
first, i will quantify the dissimilarity and then if all are fairly similar, then take mean, otherwise look closer at taxonomic assignments (maybe some tax level issues causing mismatches) and maybe drop a replicate before taking means. 


instead of filtering by the biological replicates, i need to filter by the unique extractions.  the 'dist_to_centroid' function later on doesn't like the different number of replicates or that there are so many for a few.. not sure exactly. 

how many pcr replicates does each sample have? 
```{r}
onerep <- taxon_table_2  %>%
  group_by(alternative_ID) %>%
  summarise(nrep = n_distinct(sample_id)) %>%
  #filter(nrep == 2) 
  filter(nrep == 1)
```

remove the extraction with only one pcr rep from the data frame - these are all field blanks! 
```{r}
temp_table <- taxon_table_w_meta %>%
  filter(!alternative_ID %in% onerep$alternative_ID)
```

first, i'll calculate read proportions
```{r}
normalized <- temp_table %>%
  group_by(sample_id) %>%
  mutate(Tot = sum(reads),
         Prop_reads = reads/Tot)

#add a new sample id column that also includes the location - will use this for dissimilarity measures
normalized <- normalized %>%
  unite(site_biorep, haul, alternative_ID, sep = "_", remove = FALSE) %>%
  unite(new_ID, site_biorep, sample_id, sep = "-", remove = FALSE)
```


```{r}
tibble_to_matrix <- function (tb) {
  
  tb %>%
  #normalized %>%
    group_by(new_ID, taxon) %>% 
    summarise(nReads = sum(Prop_reads)) %>% 
    spread ( key = "taxon", value = "nReads", fill = 0) %>%
    ungroup() -> matrix_1
    samples <- pull (matrix_1, new_ID)
    matrix_1[,-1] -> matrix_1
    data.matrix(matrix_1) -> matrix_1
    dimnames(matrix_1)[[1]] <- samples
    vegdist(matrix_1) -> matrix_1
}

```


```{r}
all.distances.full <- tibble_to_matrix(normalized)

# Do all samples have a name?
summary(is.na(names(all.distances.full)))
```

make the pairwise distances a long table
```{r}
as_tibble(subset(melt(as.matrix(all.distances.full)))) -> all.distances.melted

# Any major screw ups
summary(is.na(all.distances.melted$value))

# Now, create a three variables for all distances, they could be PCR replicates, BIOL replicates, or from the same site
all.distances.melted %>%
  separate (Var1, into = "Metaprobe1", sep = "\\-", remove = FALSE) %>%
  separate (Metaprobe1, into = "Haul1", remove = FALSE) %>%
  separate (Var2, into ="Metaprobe2", sep = "\\-", remove = FALSE) %>%
  separate (Metaprobe2, into = "Haul2", remove = FALSE) %>%
  mutate ( #Day.site1 = str_sub(Bottle1, start = 1, end = -2),
           #Day.site2 = str_sub(Bottle2, start = 1, end = -2),
           Distance.type = case_when( Metaprobe1 == Metaprobe2 ~ "Tech.replicates",
                                      #Day.site1 == Day.site2 ~ "Biol.replicates",
                                      Haul1 == Haul2 ~ "Same Haul",
                                      TRUE ~ "Different Haul"
                                     )) %>%
  dplyr::select(Sample1 = Var1, Sample2 = Var2 , value , Distance.type) %>%
  filter (Sample1 != Sample2) -> all.distances.to.plot

# Checking all went well
sapply(all.distances.to.plot, function(x) summary(is.na(x)))
```

```{r}
all.distances.to.plot$Distance.type <- all.distances.to.plot$Distance.type  %>% fct_relevel("Tech.replicates", "Same Haul")

ggplot (all.distances.to.plot) +
  geom_histogram (aes (fill = Distance.type, x = value, after_stat(ndensity)), position = "dodge",  alpha = 0.9, bins = 50) +
  facet_wrap( ~ Distance.type) +
  labs (x = "Pairwise dissimilarity", y = "density" ,
        Distance.type = "Distance") +
    guides (fill = "none")
```

note: eventually, it may be a good idea to separate the tech reps into extraction vs pcr replictes 

let me just look closer at the technical replicates with > 0.25 dissimilarity 
```{r}
check <- all.distances.to.plot %>%
  filter(Distance.type == "Tech.replicates") %>%
  filter(value > 0.25)

id_counts <- check %>%
  pivot_longer(
    cols = c(Sample1, Sample2),
    names_to = "which_sample",
    values_to = "ID"
  ) %>%
  count(ID, name = "n")
id_counts
```

check these hauls and specifically consider removing the above id's 
- note many are the field blanks! 

```{r}
scico_pal <- scico(24, palette = 'roma')

taxon_table_2 %>%
  filter(haul %in% c(198)) %>%
  filter(!is.na(net)) %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ggplot(aes(x=sample_id, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~net, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "sample ID") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

not going to remove any pcr reps for dissimilarity 

add a filter to remove the very very rare species - i.e. spp that never make up more that 1% of reads in any technical replicate
```{r}
#remove_rare_ids <- taxon_table_2 %>%
#  group_by(sample_id) %>%
#  mutate(total_reads = sum(reads),
#         prop_reads = reads/total_reads) %>%
#  filter(prop_reads > 0.005)
```

## summarize the technical replicates by metaprobe

fill in 0's for taxa and then calculate means 
```{r}
taxon_table_mean <- taxon_table_2 %>%
  ungroup() %>%
  select(alternative_ID, sample_type, sample_id, taxon, haul, net, depth_m, reads) %>%
  pivot_wider(names_from = taxon, values_from = reads, values_fill = 0) %>%
  pivot_longer(cols = c(7:43), names_to = "taxon", values_to = "reads") %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ungroup() %>%
  group_by(alternative_ID, sample_type, taxon, haul, net, depth_m) %>%
  summarize(mean_prop = mean(prop_reads))

taxon_table_mean %>% group_by(alternative_ID) %>% summarise(total = sum(mean_prop))
```


plot some 
```{r}
scico_pal <- scico(37, palette = 'roma')

example_hauls <- taxon_table_mean %>%
  filter(haul %in% c(196, 198, 201, 195)) %>%
  filter(!is.na(net)) %>%
  #mutate(temp_id = paste0(haul, "_", net)) %>%  # combine haul and net
  ggplot(aes(x=net, y=mean_prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~haul, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 45, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
example_hauls 

#ggsave(here("figures", "metaprobe_hauls.png"), plot = example_hauls, dpi = 300, width = 9, height = 5)
```


pivot table wide into a sample by taxon data frame - save metadata in separate df 
```{r}
wide_df <- taxon_table_mean %>% 
  filter(sample_type == "sample") %>%
  pivot_wider(names_from = "taxon", values_from = "mean_prop", values_fill = 0)
  
wide_meta <- wide_df[,c(1:5)]
wide_prop <- wide_df[,-c(1:5)]

wide_meta$var_ID = as.factor(c(1:34))
```

save wide_df to use when comparing to catch 
```{r}
#write.csv(wide_df, here("metaprobe_outputs", "taxon_wide.csv"))
```

test whether codend vs headrope differ while accounting for site-level pairings 
use bray-curtis distance matrix 
```{r}
adonis2(wide_prop ~ net,
        data = wide_meta,
        strata = wide_meta$haul,
        method = "bray")

adonis2(wide_prop ~ net,
        data = wide_meta,
        strata = wide_meta$haul,
        method = "jaccard")
```

weak evidence that community composition differ between metaprobes after accounting for haul-level pairings 
metaprobe location explains about 5% of the variation in community composition 

```{r}
# with depth
adonis2(wide_prop ~ net + depth_m,
        data   = wide_meta,
        strata = wide_meta$haul,
        method = "bray",
        by = "term")
```



but check dispersion 
```{r}
bd_bray <- betadisper(vegdist(wide_prop, "bray"), wide_meta$net)
permutest(bd_bray, strata = wide_meta$haul)

plot(bd_bray, hull = FALSE, ellipse = TRUE)

pa_table <- wide_prop
pa_table[pa_table > 0] <- 1

bd_jaccard <- betadisper(vegdist(pa_table, "jaccard"), wide_meta$net)
permutest(bd_jaccard, strata = wide_meta$haul)

plot(bd_jaccard, hull = FALSE, ellipse = TRUE)

# ggplot version
disp_site_scores <- as.data.frame(scores(bd_jaccard))
disp_site_scores$net <- wide_meta$net

ggplot(disp_site_scores, aes(x = sites.PCoA1, y = sites.PCoA2, color = net)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.68) +  # ~1 SD ellipse
  labs(x = "PCoA1", y = "PCoA2", color = "metaprobe") +
  theme_minimal()
```

permanova for bc is trustworthy since dispersion is not significant 
not the case for presence/abs - jaccard

summarize species occurrence by metaprobe location 
```{r}
species_occ <- pa_table %>%
  mutate(var_ID = rownames(pa_table)) %>%
  left_join(wide_meta %>% ungroup() %>% select(var_ID, net), by = "var_ID") %>%
  group_by(net) %>%
  summarise(across(-var_ID, sum), .groups = "drop") %>%
  pivot_longer(-net, names_to = "taxon", values_to = "count") %>%
  #pivot_wider(names_from = net, values_from = count, values_fill = 0)
  mutate(taxon_order = factor(taxon, levels = rev(unique(taxon[order(count)]))))

heatmap_plot <- ggplot(species_occ, aes(x = net, y = taxon_order, fill = count)) +
  geom_tile(color = "white") +           # makes tiles
  scale_fill_gradient(low = "white", high = "steelblue") +
  scale_y_discrete(limits = rev(levels(species_occ$taxon_order))) +  # reverse y-axis
  labs(x = "metaprobe", y = "taxa", fill = "# of observations") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8))

heatmap_plot 

#ggsave(here("figures", "metaprobe_heatmap.png"), plot = heatmap_plot, dpi = 300, width = 5, height =5)
```

simper analysis 
```{r}
sim <- simper(pa_table, wide_meta$net, permutations = 999)
summary(sim)
```


run distance-based paired test 
```{r}
dist_mat <- vegdist(wide_prop, method = "bray")

# Convert distance matrix to long form
dist_long <- as.data.frame(as.table(as.matrix(dist_mat))) %>%
  rename(sample1 = Var1,
         sample2 = Var2,
         distance = Freq) %>%
  left_join(wide_meta, by = c("sample1" = "var_ID")) %>%
  rename(haul1 = haul, net1 = net) %>%
  left_join(wide_meta, by = c("sample2" = "var_ID")) %>%
  rename(haul2 = haul, net2 = net)

paired_dist <- dist_long %>%
  filter(haul1 == haul2,
         net1 != net2) %>%
  mutate(pair_id = haul1) %>%
  distinct(pair_id, distance)

within <- paired_dist$distance
between <- dist_long %>%
  filter(haul1 != haul2,
         net1 != net2) %>%
  pull(distance)

wilcox.test(within, between)
```

again, this suggests the codend vs headrope metaprobes have similar communities 

now plot to visualize - just including net levels
```{r}
cap_net <- capscale(
  wide_prop ~ net + Condition(haul),
  data     = wide_meta,
  distance = "bray"
)
summary(cap_net)
anova(cap_net, by = "term", permutations = 999)

scores_sites <- as.data.frame(scores(cap_net, display = "sites"))
scores_sites$sample_id <- rownames(scores_sites)

scores_sites <- scores_sites %>%
  left_join(wide_meta, by = c("sample_id" = "var_ID"))

segments_df <- scores_sites %>%
  filter(net %in% c("Net_codend", "CTD_roof")) %>% 
  select(haul, net, CAP1, MDS1) %>%
  pivot_wider(
    names_from  = net,
    values_from = c(CAP1, MDS1)
  ) %>%
  drop_na()

ggplot(scores_sites, aes(x = CAP1, y = MDS1, color = net)) +
  geom_segment(
    data = segments_df,
    aes(
      x = CAP1_Net_codend, y = MDS1_Net_codend,
      xend = CAP1_CTD_roof, yend = MDS1_CTD_roof
    ),
    inherit.aes = FALSE,
    color = "grey60",
    linewidth = 0.6
  ) +
  geom_point(size = 3) +
  stat_ellipse(type = "t", level = 0.68, linewidth = 0.8) +
  labs(
    x = "dbRDA1 (net effect | haul)",
    y = "dbRDA2 (residuals)",
    color = "Net"
  ) +
  theme_bw()
```

add depth to dbRDA to get two CAP axis
```{r}
cap_net_depth <- capscale(
  wide_prop ~ net + depth_m + Condition(haul), 
  data     = wide_meta,
  distance = "bray"
)
summary(cap_net_depth)
anova(cap_net_depth, by = "term", permutations = 999)

## analogue would be: 
adonis2(wide_prop ~ net + depth_m,
        data   = wide_meta,
        strata = wide_meta$haul)

#plot 
scores_sites <- as.data.frame(scores(cap_net_depth, display = "sites"))
scores_sites$sample_id <- rownames(scores_sites)

scores_sites <- scores_sites %>%
  left_join(wide_meta, by = c("sample_id" = "var_ID"))

#lines to connect hauls
segments_df <- scores_sites %>%
  filter(net %in% c("Net_codend", "CTD_roof")) %>%   # adjust if needed
  select(haul, net, CAP1, CAP2) %>%
  pivot_wider(
    names_from  = net,
    values_from = c(CAP1, CAP2)
  ) %>%
  drop_na()

segments_df <- segments_df %>%
  mutate(
    mid_CAP1 = (CAP1_Net_codend + CAP1_CTD_roof) / 2,
    mid_CAP2 = (CAP2_Net_codend + CAP2_CTD_roof) / 2
  )

#depth vector
envfit_depth <- envfit(cap_net_depth ~ depth_m, data = wide_meta, permutations = 999)

depth_vec <- as.data.frame(scores(envfit_depth, display = "vectors"))
depth_vec$variable <- rownames(depth_vec)

# scaling factor so arrows fit nicely in the ordination space
depth_mult <- min(
  diff(range(scores_sites$CAP1)),
  diff(range(scores_sites$CAP2))
) * 0.4   # adjust 0.3–0.6 for aesthetics


depth_vec <- depth_vec %>%
  mutate(
    CAP1 = CAP1 * depth_mult,
    CAP2 = CAP2 * depth_mult
  )

# #species loadings
# sp_scores <- as.data.frame(scores(cap_net_depth, display = "species"))
# sp_scores$taxon <- rownames(sp_scores)
# 
# sp_scores <- sp_scores %>%
#   mutate(vec_len = sqrt(CAP1^2 + CAP2^2)) %>%
#   arrange(desc(vec_len)) %>%
#   slice_head(n = 5)   # adjust number
# 
# spp_mult <- min(
#   diff(range(scores_sites$CAP1)),
#   diff(range(scores_sites$CAP2))
# ) * 0.4   # adjust 0.3–0.6 for aesthetics
# 
# 
# sp_scores <- sp_scores %>%
#   mutate(
#     CAP1 = CAP1 * spp_mult,
#     CAP2 = CAP2 * spp_mult
#   )

ggplot(scores_sites, aes(x = CAP1, y = CAP2, color = net)) +
  geom_segment(
    data = segments_df,
    aes(
      x = CAP1_Net_codend, y = CAP2_Net_codend,
      xend = CAP1_CTD_roof, yend = CAP2_CTD_roof
    ),
    inherit.aes = FALSE,
    color = "grey60",
    linewidth = 0.6
  ) +
  # label segments with haul ID at midpoint
  geom_text(
    data = segments_df,
    aes(
      x = mid_CAP1,
      y = mid_CAP2,
      label = haul
    ),
    inherit.aes = FALSE,
    size = 3,
    color = "black"
  ) +
    # depth vector
  geom_segment(
    data = depth_vec,
    aes(
      x = 0, y = 0,
      xend = CAP1, yend = CAP2
    ),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")),
    color = "black",
    linewidth = 0.9
  ) +
    geom_text(
    data = depth_vec,
    aes(
      x = CAP1 * 1.2,
      y = CAP2 * 1.2,
      label = paste0("Depth")
    ),
    inherit.aes = FALSE,
    size = 3.5
  ) +
  #   # species vectors
  # geom_segment(
  #   data = sp_scores,
  #   aes(
  #     x = 0, y = 0,
  #     xend = CAP1, yend = CAP2
  #   ),
  #   inherit.aes = FALSE,
  #   arrow = arrow(length = unit(0.2, "cm")),
  #   color = "darkred",
  #   linewidth = 0.7,
  #   alpha = 0.8
  # ) +
  #  geom_text(
  #   data = sp_scores,
  #   aes(
  #     x = CAP1 * 1.2,
  #     y = CAP2 * 1.2,
  #     label = taxon
  #   ),
  #   inherit.aes = FALSE,
  #   color = "darkred",
  #   size = 3
  # ) +
  geom_point(size = 3) +
  scale_color_manual(values = c("CTD_roof" = "#007078", "Net_codend" = "#D65F00")) +
  labs(
    x = "CAP1",
    y = "CAP2",
    color = "metaprobe"
  ) +
  theme_bw()
```
