---
title: "metaprobe summary"
output: html_document
date: "2025-12-18"
---

libraries
```{r}
library(tidyverse)
library(here)
library(ggplot2)
#devtools::install_github("thomasp85/scico")
library(scico)
library(vegan)
library(reshape2)
```

load decontaminated ASV table 
```{r}
asv_table <- readRDS(here("decon_outputs", "decontaminated_asv_table.RDS"))
```

convert to taxon table
and remove any remaining pc reads 
```{r}
taxon_table <- asv_table %>%
  filter(taxon != "Acipenser fulvescens") %>% 
  group_by(extraction_ID, sample_type, sample_id, taxon, taxonomic_level, species, genus, family, order) %>%
  summarize(reads = sum(reads)) %>%
  filter(reads > 0) #%>%
  #mutate(taxon = ifelse(species == "Atheresthes evermanni", "Atheresthes evermanni", taxon)) ##should fix this in an earlier script 
```

load sample metadata 
```{r}
metadata <- read.csv(here("BeringSea2024_metaprobes_metadata.csv"))

taxon_table_w_meta <- taxon_table %>% 
  left_join(metadata)
```

make a summary table of the species and reads - dont including any field blanks in this 
```{r}
taxon_table_summary <- taxon_table_w_meta %>%
  filter(sample_type == "sample") %>%
  group_by(taxon, species, genus, family, order) %>%
  summarize(total_reads = sum(reads),
            n_tech = n_distinct(sample_id),
            n_extract = n_distinct(extraction_ID),
            n_probes = n_distinct(alternative_ID)) %>%
  arrange(desc(total_reads))

write.csv(taxon_table_summary, here("metaprobe_outputs", "taxon_table_summary.csv"), row.names = F)
```



just start by comparing the samples/pcr-reps from the same haul 
```{r}
scico_14 <- scico(14, palette = 'roma')

taxon_table_w_meta %>%
  filter(haul == 187) %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ggplot(aes(x=sample_id, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scico_14) + 
  theme_bw() +
  facet_wrap(~net, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "sample ID") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

```{r}
h187 <- taxon_table_w_meta %>%
  filter(haul == 187) %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads)
```


zoom in on some taxa - see what assignments might need updating 
```{r}
gadus <- taxon_table_w_meta %>%
  filter(genus == "Gadus") %>%
  group_by(taxon) %>%
  summarise(total_reads = sum(reads))
gadus

#prop of read to walleye pollock 
4058893/(sum(gadus$total_reads))
#prop of read to pcod
12714/(sum(gadus$total_reads))
#prop of read to gadus
127482/(sum(gadus$total_reads))

## catch weight from associated hauls was 
# 18078.545 kg pollock 
18078.545 / (18078.545+917.214)
# 917.214 kg pcod 
917.214 / (18078.545+917.214)
```
very similar pollock proportions, unassigned gadus reads likely mostly pcod...?? 

okay, next step will need to be taking the mean of the extraction/pcr replicates for samples.
first, i will quantify the dissimilarity and then if all are fairly similar, then take mean, otherwise look closer at taxonomic assignments (maybe some tax level issues causing mismatches) and maybe drop a replicate before taking means. 


instead of filtering by the biological replicates, i need to filter by the unique extractions.  the 'dist_to_centroid' function later on doesn't like the different number of replicates or that there are so many for a few.. not sure exactly. 

how many pcr replicates does each sample have? 
```{r}
onerep <- taxon_table_w_meta  %>%
  group_by(alternative_ID) %>%
  summarise(nrep = n_distinct(sample_id)) %>%
  #filter(nrep == 2) 
  filter(nrep == 1)
```

remove the extraction with only one pcr rep from the data frame - these are all field blanks! 
```{r}
temp_table <- taxon_table_w_meta %>%
  filter(!alternative_ID %in% onerep$alternative_ID)
```


first, i'll calculate read proportions
```{r}
normalized <- temp_table %>%
  group_by(sample_id) %>%
  mutate(Tot = sum(reads),
         Prop_reads = reads/Tot)

#add a new sample id column that also includes the location - will use this for dissimilarity measures
normalized <- normalized %>%
  unite(site_biorep, haul, alternative_ID, sep = "_", remove = FALSE) %>%
  unite(new_ID, site_biorep, sample_id, sep = "-", remove = FALSE)
```


```{r}
tibble_to_matrix <- function (tb) {
  
  tb %>%
  #normalized %>%
    group_by(new_ID, taxon) %>% 
    summarise(nReads = sum(Prop_reads)) %>% 
    spread ( key = "taxon", value = "nReads", fill = 0) %>%
    ungroup() -> matrix_1
    samples <- pull (matrix_1, new_ID)
    matrix_1[,-1] -> matrix_1
    data.matrix(matrix_1) -> matrix_1
    dimnames(matrix_1)[[1]] <- samples
    vegdist(matrix_1) -> matrix_1
}

```


```{r}
all.distances.full <- tibble_to_matrix(normalized)

# Do all samples have a name?
summary(is.na(names(all.distances.full)))
```

make the pairwise distances a long table
```{r}
as_tibble(subset(melt(as.matrix(all.distances.full)))) -> all.distances.melted

# Any major screw ups
summary(is.na(all.distances.melted$value))

# Now, create a three variables for all distances, they could be PCR replicates, BIOL replicates, or from the same site
all.distances.melted %>%
  separate (Var1, into = "Metaprobe1", sep = "\\-", remove = FALSE) %>%
  separate (Metaprobe1, into = "Haul1", remove = FALSE) %>%
  separate (Var2, into ="Metaprobe2", sep = "\\-", remove = FALSE) %>%
  separate (Metaprobe2, into = "Haul2", remove = FALSE) %>%
  mutate ( #Day.site1 = str_sub(Bottle1, start = 1, end = -2),
           #Day.site2 = str_sub(Bottle2, start = 1, end = -2),
           Distance.type = case_when( Metaprobe1 == Metaprobe2 ~ "Tech.replicates",
                                      #Day.site1 == Day.site2 ~ "Biol.replicates",
                                      Haul1 == Haul2 ~ "Same Haul",
                                      TRUE ~ "Different Haul"
                                     )) %>%
  dplyr::select(Sample1 = Var1, Sample2 = Var2 , value , Distance.type) %>%
  filter (Sample1 != Sample2) -> all.distances.to.plot

# Checking all went well
sapply(all.distances.to.plot, function(x) summary(is.na(x)))
```

```{r}
all.distances.to.plot$Distance.type <- all.distances.to.plot$Distance.type  %>% fct_relevel("Tech.replicates", "Same Haul")

ggplot (all.distances.to.plot) +
  geom_histogram (aes (fill = Distance.type, x = value, after_stat(ndensity)), position = "dodge",  alpha = 0.9, bins = 50) +
  facet_wrap( ~ Distance.type) +
  labs (x = "Pairwise dissimilarity", y = "density" ,
        Distance.type = "Distance") +
    guides (fill = "none")
```

let me just look closer at the technical replicates with > 0.25 dissimilarity 
```{r}
check <- all.distances.to.plot %>%
  filter(Distance.type == "Tech.replicates") %>%
  filter(value > 0.25)

id_counts <- check %>%
  pivot_longer(
    cols = c(Sample1, Sample2),
    names_to = "which_sample",
    values_to = "ID"
  ) %>%
  count(ID, name = "n")
id_counts
```

check these hauls and specifically consider removing the above id's 
- note many are the field blanks! 

```{r}
scico_pal <- scico(24, palette = 'roma')

taxon_table_w_meta %>%
  filter(haul %in% c(187, 188, 190)) %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ggplot(aes(x=sample_id, y=prop_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~net, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "sample ID") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

too high dissim. among all haul 187 ctd roof samples to include... 
188 ctd is fine 
remove e05114-B from 192 cod-end: or could update tax assignment to match others... 

look at these
```{r}
taxon_table_w_meta %>%
  filter(haul %in% c(187, 188, 190)) %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  filter(taxon %in% c("Gadus chalcogrammus", "Gadus")) %>%
  select(sample_id, taxon, prop_reads)
```


hmm.. we seem to be losing all signature of pcod. it's there but being lumped with pollock. to help with the interpretation of this study, best to lump all gadus... 


add a filter to remove the very very rare species - i.e. spp that never make up more that 1% of reads in any technical replicate
```{r}
remove_rare_ids <- taxon_table_w_meta %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  filter(prop_reads > 0.005)
```

filter results and take mean of technical replicates from the same metaprobe
```{r}
mean_taxon_table_w_meta <- remove_rare_ids %>%
  filter(alternative_ID != "202401_162_187_2") %>%
  filter(sample_id != "e05114-B") %>%
  group_by(sample_id) %>%
  mutate(total_reads = sum(reads),
         prop_reads = reads/total_reads) %>%
  ungroup() %>%
  group_by(alternative_ID, sample_type, taxon, haul, net) %>%
  summarize(mean_prop = mean(prop_reads))

mean_taxon_table_w_meta %>%
  filter(haul %in% c(187, 188, 190)) %>%
  ggplot(aes(x=alternative_ID, y=mean_prop, fill=taxon)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = scico_pal) +
  theme_bw() +
  facet_wrap(~net, scales = "free") +
  labs(
    y = "proportion of sequencing reads",
    x = "sample ID") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    #legend.position = "none",
    legend.title = element_blank()
  )
```

pivot table wide into a sample by taxon dataframe - save metadata in seperate df 
```{r}
wide_df <- mean_taxon_table_w_meta %>% 
  pivot_wider(names_from = "taxon", values_from = "mean_prop", values_fill = 0) %>%
  filter(sample_type == "sample")

wide_meta <- wide_df[,c(1:4)]
wide_prop <- wide_df[,-c(1:4)] %>% select(!`Gadus macrocephalus`) ## no reads for pcod 
```

note: the Gadus is likely pollock..? or look at catch and decide if i need to call all chalco and macro reads gadus... 

use bray-curtis distance matrix 
```{r}
dist <- vegdist(wide_prop, method = "bray")

as_tibble(subset(melt(as.matrix(dist)))) -> distances.melted

wide_meta$VarID = c(1:33)

dist_df <- distances.melted %>%
  left_join(wide_meta, by = c("Var1" = "VarID")) %>%
  select(haul, net, value, Var2) %>%
  rename(haul1 = haul) %>%
  rename(net1 = net) %>%
  mutate(id1 = paste(haul1, net1, sep = "_")) %>%
  left_join(wide_meta, by = c("Var2" = "VarID")) %>%
  select(haul1, net1, id1, value, haul, net) %>%
  rename(haul2 = haul) %>%
  rename(net2 = net) %>%
  mutate(id2 = paste(haul2, net2, sep = "_")) %>%
  filter(id1 != id2) %>%
  mutate(pair_type = if_else(haul1 == haul2,
                             "within_haul",
                             "between_haul"))

ggplot (dist_df) +
  geom_histogram (aes (fill = pair_type, x = value, after_stat(ndensity)), position = "dodge",  alpha = 0.9, bins = 50) +
  facet_wrap( ~ pair_type) +
  labs (x = "pairwise dissimilarity", y = "density" ,
        Distance.type = "Distance") +
    guides (fill = "none")
```




