---
title: "metaprobes vs catch"
output: html_document
date: "2025-12-30"
---

libraries
```{r}
library(tidyverse)
library(here)
library(ggplot2)
#devtools::install_github("thomasp85/scico")
library(scico)
library(vegan)
library(reshape2)
```

read in file to join/merge nets and metaprobes
```{r}
key <- read.csv(here("tax_join.csv"))

key_unique <- key %>%
  group_by(eDNA_taxa) %>%
  slice(1) %>%  # take the first row if duplicates exist
  ungroup()

key_unique_catch <- key %>%
  group_by(catch_species_name) %>%
  slice(1) %>%  # take the first row if duplicates exist
  ungroup()

key_curated <- key %>%
  select(curated, curated_common) %>%
  unique()
key_curated
```

read in formated metaprobe data
```{r}
metaprobe_df <- read.csv(here("metaprobe_outputs", "taxon_wide.csv")) %>% select(!X) %>%
  pivot_longer(cols = -c(1:5), names_to = "eDNA_taxa", values_to = "proportion") %>%
  mutate(eDNA_taxa = gsub("\\.", " ", eDNA_taxa)) %>%
  left_join(key_unique, by = "eDNA_taxa") %>%
  group_by(haul, net, depth_m, curated, curated_common) %>%
  summarise(proportion = sum(proportion)) %>%
  mutate(net = ifelse(net == "Net_codend", "codend", net),
         net = ifelse(net == "CTD_roof", "roof", net))
```

read in catch data
```{r}
catch <- read.csv(here("catch_outputs", "metaprobe_haul_catch.csv")) %>%
  select(weight_kg, species_name, haul, depth_m) %>%
  left_join(key_unique_catch, by = c("species_name" = "catch_species_name")) %>%
  group_by(haul, depth_m, curated, curated_common) %>%
  summarize(weight_kg = sum(weight_kg)) %>%
  group_by(haul) %>%
  mutate(total_kg = sum(weight_kg),
         proportion = weight_kg/total_kg) %>%
  mutate(net = "trawl") %>%
  select(haul, net, depth_m, curated, curated_common, proportion)
```

combine
```{r}
DF <- metaprobe_df %>%
  bind_rows(catch)
```

pivot table wide into a sample by taxon data frame - save metadata in separate df 
```{r}
DF_wide <- DF %>%
  ungroup() %>%
  select(!curated) %>%
  pivot_wider(names_from = "curated_common", values_from = "proportion", values_fill = 0)

DF_meta <- DF_wide[,c(1:3)]
DF_prop <- DF_wide[,-c(1:3)]
```

okay, first explore the data with an NMDS 
```{r}
# Bray–Curtis (abundance / proportions)
dist_bray <- vegdist(DF_prop, method = "bray")

# Jaccard (presence–absence)
pa_mat <- DF_prop
pa_mat[pa_mat > 0] <- 1
dist_jaccard <- vegdist(pa_mat, method = "jaccard")

set.seed(123)

#run NMDS
nmds_bray <- metaMDS(
  dist_bray,
  k = 2,
  trymax = 200,
  autotransform = FALSE
)

nmds_jaccard <- metaMDS(
  dist_jaccard,
  k = 2,
  trymax = 200,
  autotransform = FALSE
)

#check stress
nmds_bray$stress
nmds_jaccard$stress

#extract site scores
scores_bray <- scores(nmds_bray, display = "sites") %>%
  as.data.frame() %>%
  bind_cols(DF_meta)

scores_jaccard <- scores(nmds_jaccard, display = "sites") %>%
  as.data.frame() %>%
  bind_cols(DF_meta)
```

plot bray-curtis
```{r}
#plot 
ggplot(scores_bray, aes(NMDS1, NMDS2, color = net)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = net), linewidth = 1) +
  theme_bw() +
  labs(
    title = "NMDS – Bray–Curtis",
    color = "Metaprobe"
  )

scico_pal <- scico(17, palette = 'imola')

nmds_bray <- ggplot(scores_bray, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = as.factor(haul), shape = as.factor(net)), size = 3) +
  stat_ellipse(aes(group = net, linetype = net), linewidth = 1) +
  scale_color_manual(
    values = scico_pal,
    name = "Haul"
  ) +
  theme_bw() +
  labs(
    title = "NMDS – Bray–Curtis",
    color = "Haul",
    shape = "Gear",
    linetype = "Gear"
  )
nmds_bray
#ggsave(here("figures", "nmds_bray.png"), plot = nmds_bray, dpi = 300, width = 8, height = 6)

nmds_bray_facet <-ggplot(scores_bray, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = as.factor(haul), shape = net), size = 3) +
  scale_color_manual(
    values = scico_pal,
    name = "Haul"
  ) +
  geom_line(
    aes(group = haul),
    color = "grey60",
    linewidth = 0.7
  ) +
  facet_wrap(~ haul) +
  theme_bw() +
  guides(color = "none") +
  labs(
    title = "NMDS (Bray-Curtis)",
    shape = "gear"
  )
nmds_bray_facet
#ggsave(here("figures", "nmds_bray_facet.png"), plot = nmds_bray_facet, dpi = 300, width = 8, height = 6)
```


plot jaccard
```{r}
ggplot(scores_jaccard, aes(NMDS1, NMDS2, color = net)) +
  geom_point(size = 3) +
  stat_ellipse(aes(group = net), linewidth = 1) +
  theme_bw() +
  labs(
    title = "NMDS – Jaccard (presence–absence)"
  )

ggplot(scores_jaccard, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = net), size = 3) +
  geom_line(
    aes(group = haul),
    color = "grey60",
    linewidth = 0.7
  ) +
  facet_wrap(~ haul) +
  theme_bw() +
  labs(
    title = "Paired NMDS (Jaccard)",
    color = "Metaprobe"
  )

scico_pal <- scico(17, palette = 'imola')

nmds_jaccard <- ggplot(scores_jaccard, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = as.factor(haul), shape = as.factor(net)), size = 3) +
  stat_ellipse(aes(group = net, linetype = net), linewidth = 1) +
  scale_color_manual(
    values = scico_pal,
    name = "Haul"
  ) +
  theme_bw() +
  labs(
    title = "NMDS – Jaccard",
    color = "Haul",
    shape = "Gear",
    linetype = "Gear"
  )
#ggsave(here("figures", "nmds_jaccard.png"), plot = nmds_jaccard, dpi = 300, width = 8, height = 6)
```


test whether codend vs headrope vs trawl differ while accounting for site-level pairings - PERMANOVA
use bray-curtis and jaccard distance matrices
```{r}
adonis2(DF_prop ~ net,
        data = DF_meta,
        strata = DF_meta$haul,
        method = "bray")

adonis2(DF_prop ~ net,
        data = DF_meta,
        strata = DF_meta$haul,
        method = "jaccard")
```

but check dispersion 
```{r}
bd_bray <- betadisper(vegdist(DF_prop, "bray"), DF_meta$net)
permutest(bd_bray, strata = DF_meta$haul)

plot(bd_bray, hull = FALSE, ellipse = TRUE)

pa_table <- DF_prop
pa_table[pa_table > 0] <- 1

bd_jaccard <- betadisper(vegdist(pa_table, "jaccard"), DF_meta$net)
permutest(bd_jaccard, strata = DF_meta$haul)

plot(bd_jaccard, hull = FALSE, ellipse = TRUE)

# ggplot version
#disp_site_scores <- as.data.frame(scores(bd_jaccard))
disp_site_scores <- as.data.frame(scores(bd_bray))
disp_site_scores$net <- DF_meta$net

ggplot(disp_site_scores, aes(x = sites.PCoA1, y = sites.PCoA2, color = net)) +
  geom_point(size = 3) +
  stat_ellipse(level = 0.68) +  # ~1 SD ellipse
  labs(x = "PCoA1", y = "PCoA2", color = "metaprobe") +
  theme_minimal()
```


now use a dbRDA
```{r}
dbrda_net_only <- capscale(
  DF_prop ~ net + Condition(haul),
  data     = DF_meta,
  distance = "bray"
)

dbrda_net <- capscale(
  DF_prop ~ net + depth_m + Condition(haul),
  data     = DF_meta,
  distance = "bray"
)

dbrda_net_only
dbrda_net
anova(dbrda_net_only, permutations = 999)
anova(dbrda_net, permutations = 999)
anova(dbrda_net, by = "term", permutations = 999)
```

plot dbRDA - with depth 
```{r}
# site scores
scores_sites <- scores(dbrda_net, display = "sites") %>% 
  as.data.frame() %>% 
  bind_cols(DF_meta)

# eigenvalues
cap_eig <- dbrda_net$CCA$eig  # vector of CAP1, CAP2, CAP3
percent_var_CAP <- round(100 * cap_eig / sum(cap_eig), 1)
percent_var_total <- round(100 * cap_eig / sum(dbrda_net$tot.chi), 1)

# vector for depth
depth_vec <- envfit(dbrda_net, DF_meta["depth_m"], permutations = 999)
depth_scores <- scores(depth_vec, display = "vectors") %>%
  as.data.frame() %>%
  rownames_to_column("variable")

library(ggplot2)
library(grid)  # for arrow()

cap_bray <- ggplot(scores_sites, aes(x = CAP1, y = CAP2)) +
  
  # ellipses for gear
  stat_ellipse(aes(group = net, fill = net),
               geom = "polygon",
               alpha = 0.25,
               color = NA) + 
  
  scale_fill_manual(values = c("roof" = "#007078", "codend" = "#D65F00", "trawl" = "#625BC4")) +

  # points by net/gear
  geom_point(aes(shape = net, color = as.factor(haul)), size = 3) +

  # depth vector
  geom_segment(
    data = depth_scores,
    aes(
      x = 0, y = 0,
      xend = CAP1, yend = CAP2
    ),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")),
    color = "black",
    linewidth = 0.9
  ) +
  geom_text(
    data = depth_scores,
    aes(
      x = CAP1 * 1.4,
      y = CAP2 * 1.4,
      label = variable
    ),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  # optional: customize colors
  scale_color_manual(values = scico_pal) +

  # axis labels with % variance
  labs(
    x = paste0("CAP1 (", percent_var_total[1], "%)"),
    y = paste0("CAP2 (", percent_var_total[2], "%)"),
    shape = "Gear",
    color = "Haul",
    fill = "Gear"
  ) +
  guides(color = "none") +
  theme_bw()

cap_bray 
#ggsave(here("figures", "cap_bray.png"), plot = cap_bray, dpi = 300, width = 8, height = 6)
```


plot dbRDA - without depth 
```{r}
# site scores
scores_sites <- scores(dbrda_net_only, display = "sites") %>% 
  as.data.frame() %>% 
  bind_cols(DF_meta)

# eigenvalues
cap_eig <- dbrda_net_only$CCA$eig  # vector of CAP1, CAP2
percent_var_CAP <- round(100 * cap_eig / sum(cap_eig), 1)
percent_var_total <- round(100 * cap_eig / sum(dbrda_net$tot.chi), 1)

library(ggplot2)
library(grid)  # for arrow()

ggplot(scores_sites, aes(x = CAP1, y = CAP2)) +
  
  # ellipses for gear
  stat_ellipse(aes(group = net, fill = net),
               geom = "polygon",
               alpha = 0.25,
               color = NA) + 
  
  scale_fill_manual(values = c("roof" = "#007078", "codend" = "#D65F00", "trawl" = "#625BC4")) +
  
  # points by net/gear
  geom_point(aes(shape = net, color = as.factor(haul)), size = 3) +
  
  # optional: customize colors
  scale_color_manual(values = scico_pal) +

  # axis labels with % variance
  labs(
    x = paste0("CAP1 (", percent_var_total[1], "%)"),
    y = paste0("CAP2 (", percent_var_total[2], "%)"),
    shape = "Gear",
    color = "Haul",
    fill = "Gear"
  ) +
  guides(color = "none") +
  theme_bw()
```


dbRDA - jaccard
```{r}
dbrda_net_only <- capscale(
  pa_table ~ net + Condition(haul),
  data     = DF_meta,
  distance = "jaccard"
)

dbrda_net <- capscale(
  pa_table ~ net + depth_m + Condition(haul),
  data     = DF_meta,
  distance = "bray"
)

dbrda_net_only
dbrda_net
anova(dbrda_net_only, permutations = 999)
anova(dbrda_net, permutations = 999)
anova(dbrda_net, by = "term", permutations = 999)
```

plot dbRDA - with depth 
```{r}
# site scores
scores_sites <- scores(dbrda_net, display = "sites") %>% 
  as.data.frame() %>% 
  bind_cols(DF_meta)

# eigenvalues
cap_eig <- dbrda_net$CCA$eig  # vector of CAP1, CAP2, CAP3
percent_var_CAP <- round(100 * cap_eig / sum(cap_eig), 1)
percent_var_total <- round(100 * cap_eig / sum(dbrda_net$tot.chi), 1)

# vector for depth
depth_vec <- envfit(dbrda_net, DF_meta["depth_m"], permutations = 999)
depth_scores <- scores(depth_vec, display = "vectors") %>%
  as.data.frame() %>%
  rownames_to_column("variable")

library(ggplot2)
library(grid)  # for arrow()

cap_jaccard <- ggplot(scores_sites, aes(x = CAP1, y = CAP2)) +
  
  # ellipses for gear
  stat_ellipse(aes(group = net, fill = net),
               geom = "polygon",
               alpha = 0.25,
               color = NA) + 
  
  scale_fill_manual(values = c("roof" = "#007078", "codend" = "#D65F00", "trawl" = "#625BC4")) +

  # points by net/gear
  geom_point(aes(shape = net, color = as.factor(haul)), size = 3) +

  # depth vector
  geom_segment(
    data = depth_scores,
    aes(
      x = 0, y = 0,
      xend = CAP1, yend = CAP2
    ),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.25, "cm")),
    color = "black",
    linewidth = 0.9
  ) +
  geom_text(
    data = depth_scores,
    aes(
      x = CAP1 * 1.4,
      y = CAP2 * 1.4,
      label = variable
    ),
    inherit.aes = FALSE,
    size = 3.5
  ) +

  # optional: customize colors
  scale_color_manual(values = scico_pal) +

  # axis labels with % variance
  labs(
    x = paste0("CAP1 (", percent_var_total[1], "%)"),
    y = paste0("CAP2 (", percent_var_total[2], "%)"),
    shape = "Gear",
    color = "Haul",
    fill = "Gear"
  ) +
  guides(color = "none") +
  theme_bw()

cap_jaccard 
#ggsave(here("figures", "cap_jaccard.png"), plot = cap_jaccard, dpi = 300, width = 8, height = 6)
```

heatmap of occurance by gear 
```{r}
DF_meta$var_ID = as.character(c(1:51))

species_occ <- pa_table %>%
  mutate(var_ID = rownames(pa_table)) %>%
  left_join(DF_meta %>% ungroup() %>% select(var_ID, net), by = "var_ID") %>%
  group_by(net) %>%
  summarise(across(-var_ID, sum), .groups = "drop") %>%
  pivot_longer(-net, names_to = "taxon", values_to = "count") %>%
  #pivot_wider(names_from = net, values_from = count, values_fill = 0)
  mutate(taxon_order = factor(taxon, levels = rev(unique(taxon[order(count)]))))

heatmap_gear <- ggplot(species_occ, aes(x = net, y = taxon_order, fill = count)) +
  geom_tile(color = "white") +           # makes tiles
  scale_fill_gradient(low = "white", high = "steelblue") +
  scale_y_discrete(limits = rev(levels(species_occ$taxon_order))) +  # reverse y-axis
  labs(x = "gear", y = "taxa", fill = "# of observations") +
  theme_bw() +
  theme(axis.text.y = element_text(size = 8))

heatmap_gear 
#ggsave(here("figures", "gear_heatmap.png"), plot = heatmap_gear, dpi = 300, width = 6, height =5)
```

plot some hauls by gear 
```{r}
scico_pal <- scico(31, palette = 'roma')

exampleB <- DF %>%
  #filter(haul %in% c(192, 196, 201)) %>%
  filter(haul %in% c(189, 195, 198)) %>%
  filter(proportion > 0) %>%
  ggplot(aes(x=net, y=proportion, fill=curated_common)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = scico_pal) + 
  theme_bw() +
  facet_wrap(~haul, scales = "free") +
  labs(
    y = "proportion",
    x = "") + 
  theme(
    axis.text.x = element_text(angle = 34, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "right",
    legend.title = element_blank()
  )

exampleB 
#ggsave(here("figures", "comp_by_gearB.png"), plot = exampleB, dpi = 300, width = 8, height =4)
```



